---
title: "Download the airlines data"
subtitle: "USCOTS 2025 breakout session"
author: "Nicholas Horton (nhorton@amherst.edu) and Jo Hardin (jo.hardin@pomona.edu)"
date: "2025-07-18"
date-format: long
format: html
execute:
  warning: false
  message: false
toc: true
editor: source
---

```{r}
#| label: setup
#| include: false
library(tidyverse)
library(duckdb)
library(tictoc)
rnd_val <- 2
```

The current Quarto file downloads the airline flight data from the American Statistical Association's Data Expo 2024.
The data are stored in [Parquet format](https://www.databricks.com/glossary/what-is-parquet#:~:text=Apache%20Parquet%20is%20an%20open,handle%20complex%20data%20in%20bulk.), which is efficient for large datasets and can be read using SQL commands using the **arrow** package in R.

Once you've successfully rendered the current Quarto file (`1_download_data.qmd`), open and render `2_explore_sql.qmd` to carry out an analysis of the downloaded data.
(Note that both files must be saved and rendered from the same folder to access the downloaded data.)

See https://community.amstat.org/dataexpo/home for background and the source for much of the code below and https://github.com/nicholasjhorton/uscots2025-sql-data-technologies for more resources.

```{r}
tictoc::tic()

folder_name <- "data_airlines"
dir.create(folder_name, showWarnings = TRUE)
stopifnot(file.exists(folder_name))

list.files(folder_name)
base_url <- "https://blobs.duckdb.org/flight-data-partitioned/"

years <- 2023:2024 # could be 1987:2024, but that's a *lot* of data!
# please don't download too much data during the breakout session

files <- paste0("Year=", years, "/data_0.parquet")
my_files <- paste0(folder_name, "/Year=", years, "/data_0.parquet")

for (dir in dirname(my_files)) {
  dir.create(dir, showWarnings = FALSE)
}

out <- curl::multi_download(paste0(base_url, files), my_files, resume = TRUE)

tictoc::toc()
```

The **tictoc** package will tell you how long it took to download the files into your local system.
It should not take too long since you are only downloading a proper subset of what's there (to avoid overloading the network at the breakout session).

```{r}
list.files(folder_name)
```

We will be accessing the parquet files in the next Quarto file, `2_explore_sql.qmd`.
